Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2018-01-25 writeup/conStruct.tex
added Graham edits

A classic problem in population genetics is the characterization 
of discrete population structure in the presence of 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
the use of clustering or assignment methods may incorrectly ascribe
differentiation due to continuous processes (e.g., geographic isolation by distance)
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
This reflects a shortcoming of current methods for inferring and 
visualizing population structure when applied to genetic data
deriving from geographically distributed populations.
Here, we present a statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of two-dimensional population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This thereby explicitly addresses the ``clines versus clusters" problem in 
modeling population genetic variation,
and remedies the overfitting that nonspatial models are prone to.
The method produces useful descriptions of structure in genetic relatedness
in situations where separated, geographically distributed populations interact,
as after a range expansion or secondary contact.
We demonstrate the utility of this approach using simulations 
and by applying it to empirical datasets of poplars and black bears in North America.

@@

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \cite{CavalliSforza1975, treemix},
dimensionality-reduction approaches, 
such as principal components analysis 
\cite{meirmans2009genodive,menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
(e.g., \cite{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}), among others.
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of differentiation,
as that produced by geographic isolation by distance 
\cite{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on ancestry proportions \cite{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.

@@

\paragraph{Data}
The statistical framework of our approach is conceptually similar to that in \cite{Wasser2004}, \cite{BEDASSLE}, and \cite{spacemix},
although we use a somewhat different summary statistic than in this previous work.
The model works with allele frequencies at $L$ unlinked, bi-allelic single nucleotide polymorphisms (SNPs) genotyped across $N$ samples.
Each ``sample'' may be a single individual,  
a collection of individuals from a location, 
or frequencies estimated from pooled sequencing.
From these we compute the \emph{allelic covariance} 
between samples $i$ and $j$, denoted $\widehat{\Omega}_{i,j}$,
as the expected covariance of distinct individual alleles 
chosen from each of the two samples at a random locus.
More precisely, suppose that we pick a random bi-allelic locus uniformly,
pick a random ``reference'' allelic state from the two seen at that locus,
and, in each sample, draw one random allele,
recording $X_i = 1$ if the allele drawn in sample $i$ matches the random reference,
and $X_i = 0$ otherwise.
Then, 
\begin{equation} \label{eqn:first_allelic_cov}
    \widehat{\Omega}_{i,j} = \cov[X_i, X_j] .
\end{equation}
Each $X_i$ behaves marginally as a fair coin --
in particular, $\P\{X_i = 1\} = 1/2$, so $\widehat{\Omega}_{i,i} = 1/4$ for every $i$ --
all information enters through \emph{correlations}.

@@

If the genomic coordinates of the loci are known, 
the training/testing partitioning should be designed to accommodate linkage disequilibrium (LD).
Loci in strong LD are not inherited independently,
so if loci from a single linkage block are included in both training and testing partitions, 
the independence of the test in the testing partition will be compromised,
and so the parameters estimated from the training partition 
might be describing process heterogeneity or noise 
in a region of the genome that also has loci included in the testing partition.
The best practice for cross-validation is to make sure that 
no loci in the testing dataset are in strong LD with, 
or near to on the genome, loci in the testing dataset.

>>>>

A classic problem in population genetics is the characterization 
of discrete population structure in the presence of 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
the use of clustering or assignment methods may incorrectly ascribe
differentiation due to continuous processes (e.g., geographic isolation by distance)
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
This reflects a shortcoming of current methods for inferring and 
visualizing population structure when applied to genetic data
deriving from geographically distributed populations.
Here, we present a statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of two-dimensional population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This thereby explicitly addresses the ``clines versus clusters" problem in 
modeling population genetic variation,
and remedies some of the overfitting that nonspatial models are prone to.
The method produces useful descriptions of structure in genetic relatedness
in situations where separated, geographically distributed populations interact,
as after a range expansion or secondary contact.
We demonstrate the utility of this approach using simulations 
and by applying it to empirical datasets of poplars and black bears in North America.

@@

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \cite{CavalliSforza1975, treemix},
dimensionality-reduction approaches, 
such as principal components analysis 
\cite{meirmans2009genodive,menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
(e.g., \cite{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}), among others.
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of differentiation,
as that produced by geographic isolation by distance 
\cite{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on cluster membership \cite{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.

@@

\paragraph{Data}
The statistical framework of our approach is conceptually similar to that in \cite{Wasser2004}, \cite{BEDASSLE}, and \cite{spacemix},
although we use a somewhat different summary statistic than in this previous work.
The model works with allele frequencies at $L$ unlinked, bi-allelic single nucleotide polymorphisms (SNPs) genotyped across $N$ samples.
Each ``sample'' may be a single individual,  
a collection of individuals from a location, 
or frequencies estimated from pooled sequencing.
From these we compute the \emph{allelic covariance} 
between samples $i$ and $j$, denoted $\widehat{\Omega}_{i,j}$,
as the expected covariance of distinct individual alleles 
chosen from each of the two samples at a random locus.
More precisely, suppose that we pick a random bi-allelic locus uniformly,
pick a random ``reference'' allelic state from the two alleles seen at that locus,
and, in each sample, draw one random allele,
recording $X_i = 1$ if the allele drawn in sample $i$ matches the random reference,
and $X_i = 0$ otherwise.
Then, 
\begin{equation} \label{eqn:first_allelic_cov}
    \widehat{\Omega}_{i,j} = \cov[X_i, X_j] .
\end{equation}
Because we randomly choose the reference allele, 
each $X_i$ behaves marginally as a fair coin --
in particular, $\P\{X_i = 1\} = 1/2$, so $\widehat{\Omega}_{i,i} = 1/4$ for every $i$ --
all information enters through \emph{correlations}.

@@

\gb{If the spatial coordinates of the loci in the genome are known, 
the training/testing partitioning should be designed to accommodate linkage disequilibrium (LD).
Loci in strong LD do not represent independent instantiations of the coalescent process, 
so if loci from a single linkage block are randomly split into both training and testing partitions, 
the independence of the test in the testing partition will be compromised 
(e.g., the parameters estimated from the training partition 
might be describing process heterogeneity or noise 
in a region of the genome that also has loci included in the testing partition).
If genomic position or LD between SNPs is known, 
the best practice for cross-validation is to make sure that 
no loci in the testing dataset are in strong LD with, 
or spatially proximate to, loci in the testing dataset.}
